Parallel Level-Set Methods on Adaptive Tree-Based Grids

### 2016-02-14 paper/peer-reviews/response_letter.tex

\documentclass{scrlttr2}

\begin{document}
\begin{letter} {Editorial Office, \\ Journal of Computational Physics}
\setkomavar{subject}{Re: JCOMP-D-15-00846}
\opening{Dear Editor,}
We would like to appreciate the invaluable comments made by the reviewers. We have made adjustments to our article, titled `\textbf{Parallel level-set methods on adaptive tree-based grids}', to address the issues raised by the reviewers as detailed below.

\begin{itemize}
\item \textbf{Reviewer \#1} \\
The first reviewer has suggested to include ``\textit{some basic library codes in appendix, each matching its level set algorithm}''. Indeed it is important for the scientific community to share knowledge and facilitate ways for reproducing scientific work for others. In the case of computational science, unfortunately, this is a very daunting task that is not always possible. Many scientific codes are highly specialized and are often too technical to be fitted into a paper.

Specifically, in our case, our library amounts to nearly 30,000 lines of code, spread across more than 50 files and with complex dependencies. As such, it is practically not viable to include sample codes in the paper. Moreover, we believe inclusion of actual code, which involves many technical details, e.g. error checking, profiling tools, comments, etc., is not scientifically educational and may further alienate the reader from main message of the paper.

Instead, and in accordance to the second reviewer who has raised related issues, we have opted to revise the pseudo-codes in the algorithms to make them more clear. We believe that the modifications should make the algorithms easier to understand and useful to the readers.

\item \textbf{Reviewer \#2} \\
The second reviewer has raised several issues, mostly related to the pseudo-codes in the paper, as well as suggesting additional numerical tests. These issues are addressed as detailed below:
\begin{itemize}
\item ``\textit{It is recommended that the authors' pseudocode be closer to a language that many people are familiar with like C++ or Fortran 90}.''

Our library is coded in `\texttt{c++}' (with a mix of little `\texttt{c}' code). We have made adjustments to the to algorithms to make them easier to understand. However, we believe it is best not to directly use a `\texttt{c++}' syntax to keep the pseudo-code language-agnostic and accessible to all readers.

\item ``\textit{In Algorithm 2, what is the definition of “rank”? “owner’s rank”? “mpirank”? “st”?}''

\item ``\textit{}''

\item ``\textit{}''

\item ``\textit{}''
\end{itemize}
\end{itemize}

In light of these adjustments, we respectfully ask that our article be reconsidered for publication in the Journal of Computational Physics.
\closing{Sincerely,}
Mohammad Mirzadeh (on behalf of all authors)
\end{letter}
\end{document}

### 2016-02-15 paper/tex/parallel_algorithms.tex
Made changes to paper and prepared the response letter

Finally, in \texttt{p4est} trees are linearized, i.e.\ only the leaves are explicitly stored. However, explicit knowledge of the hierarchal structure of the tree is greatly beneficial in several algorithms, e.g. in search operations needed for the interpolation algorithm. Thus, we introduce a simple reconstruction algorithm that recreates a local representation of the entire ``forest'' that is only adapted to local cells and, potentially, the ghost layer. This approach is similar to the ideas introduced in \cite{Bangerth;Burstedde;Heister;etal:11:Algorithms-and-data-} and our tests show that in a typical application they amount to less that $1\%$ of the entire runtime. Algorithm \ref{alg:reconstruction} illustrates how this reconstruction is performed. Given a forest and a layer of ghost cells from \texttt{p4est}, the algorithm generates a local representation of the forest by recursively refining from the root until reaching the same level and location of all leaves in the local forest and the ghost layer. Note that algorithm \ref{alg:reconstruction} does not involve any communication and is load balanced provided that the initial forest is balanced. Figure \ref{fig:reconstruction} illustrates an example where Algorithm \ref{alg:reconstruction} is applied. Note how each process has independently generated a local representation of the forest that is refined to match the same leaves as in the global forest and ghost layer.
\begin{figure}[htbp]
\begin{center}
\includegraphics[width = \textwidth]{figures/reconstruct.pdf}
\end{center}
\caption{Left: a forest refined close to an interface and partitioned among four processes, as indicated by colors. Right: each process independently recreates a local forest that is refined to match the local grid and is as coarse as possible elsewhere. Note that empty cells are fictitious, i.e.\ they are only required to generate the hierarchal structure and are not matched by any corresponding cell in the global forest.}
\label{fig:reconstruction}
\end{figure}
\begin{algorithm}[htbp]
\caption{$H \gets \texttt{Reconstruct (}G\texttt{)}$: construction of the local tree hierarchy $H$ from the grid $G$}
\begin{algorithmic}[1]
\State $H \gets G.\texttt{macromesh()}$
\For {$\mathit{tr} : G.\texttt{local\_trees()}$} \Comment{build hierarchy for local cells}
	\For {$c : \mathit{tr}.\texttt{cells()}$}
		\State $H.\texttt{update\_tree(} \mathit{tr}, c \texttt{)}$
	\EndFor
\EndFor

\For {$c : G.\texttt{ghost\_cells()}$} \Comment{build hierarchy for ghost cells}
	\State $H.\texttt{update\_tree(} c.\texttt{tree()}, c \texttt{)}$
\EndFor

\State \Return $H$
\\
\Function{$H.\texttt{update\_tree}$}{$\mathit{tr}, c$}   \Comment{recursive tree reconstruction}
	\State $c_l \gets H.\texttt{root(}\mathit{tr}\texttt{)}$
	\While{$c_l.\texttt{level()} \not= c.\texttt{level()} $}
		\If {$c_l.$\texttt{is\_leaf()}} $c_l$\texttt{.split()}
		\EndIf
		\State $h \gets c_l.\texttt{length()} / 2$ \Comment{select the next child based on direction}
		\State $i \gets c.x \ge c_l.x + h$
		\State $j \gets c.y \ge c_l.y + h$
		\State $k \gets c.z \ge c_l.z + h$

		\State $c_l \gets c_l.\texttt{child(} i, j, k \texttt{)}$
	\EndWhile	
\EndFunction
\end{algorithmic}
\label{alg:reconstruction}
\end{algorithm}

\subsection{Interpolation and semi-Lagrangian methods}
As indicated earlier, we use the semi-Lagrangian method to solve equation \eqref{eq:ls} when the velocity field is externally generated, i.e.\ when it does not depend explicitly on the level-set function itself. Let us rewrite equation \eqref{eq:ls} along the characteristic curve $\underline{\mathbf{X}}(t)$ as:
\be
\left\{
\begin{array}{rcl}
\dfrac{\text{d} \underline{\mathbf{X}}}{\text{d} t} &=& \underline{\mbf{u}}, \\ [3ex]
\dfrac{\text{d} \phi(\underline{\mathbf{X}}(t), t)}{\text{d} t} &=& 0.
\end{array}
\right.
\label{eq:sl}
\ee
The semi-Lagrangian method integrates equations \eqref{eq:sl} backward in time,
i.e.\ starting from the grid $G^{n+1}$ (computed in iteratively as explained
later on), we simply write $\phi^{n+1}(\underline{\mathbf{X}}^{n+1}) =
\phi(\underline{\mathbf{X}}(t^{n+1}), t^{n+1}) =
\phi(\underline{\mathbf{X}}(t^n), t^n) = \phi^n(\underline{\mathbf{X}}_d)$.
Here, the characteristic curves are chosen such that $\underline{\mathbf{X}}(t^{n+1})$ are simply the coordinates of grids points of $G^{n+1}$, and $\underline{\mathbf{X}}_d$ are the departure points, which are computed using the second-order midpoint method \cite{Min;Gibou:07:A-second-order-accur}:
\bea
\underline{\mathbf{X}}^\star &=& \underline{\mathbf{X}}^{n+1} - \frac{\Delta t}{2} \underline{\mbf{u}}^{n}(\underline{\mathbf{X}}^n),	   \label{eq:xstar}      \\
\underline{\mathbf{X}}_d     &=& \underline{\mathbf{X}}^{n+1} - \Delta t \underline{\mbf{u}}^{n+\frac{1}{2}}(\underline{\mathbf{X}}^\star), \label{eq:xdeparture}
\eea
where $\underline{\mbf{u}}^{n+\frac{1}{2}}$ is obtained via extrapolation from previous times, i.e.:
\be
\underline{\mbf{u}}^{n+\frac{1}{2}} = \frac{3}{2} \underline{\mbf{u}}^n - \frac{1}{2}\underline{\mbf{u}}^{n-1}. \label{eq:vn_p_half}
\ee
Note that all values at the intermediate point, $\underline{\mathbf{X}}^\star$,
and departure point, $\underline{\mathbf{X}}_d$, must be calculated via
interpolation from the previous grids $G^{n}$ and $G^{n-1}$.
Here, we use the stabilized second-order interpolation for
$\phi(\underline{\mathbf{X}}_d)$ and the multi-linear interpolation for
$\underline{\mbf{u}}^{n+\frac{1}{2}}(\underline{\mathbf{X}}^\star)$
\cite{Min;Gibou:07:A-second-order-accur}.
Although parallelization of the interpolation process on a shared-memory
machine is trivial, the same cannot be said for distributed-memory machines.
In fact, the parallel interpolation given in Algorithm \ref{alg:interpolation}
is probably the most important contribution of this article since this
procedure, which is trivial on uniform grids, is challenging in the case of
trees because it is not straightforward to identify which processes owns the
departure points.
Indeed, complications arise because not all departure points will reside in the
domain owned by the current process. Moreover, due to the irregular shapes of
the partitions, one cannot even ensure they are entirely owned by neighboring
processes.
At best we can only expect that their locations are bounded by a halo of width
$w \le \text{CFL} \: \Delta x_{\min}$ around the local partition, where
$x_{\min}$ is the size of the smallest cell in the forest.
Naturally, if one enforces $\text{CFL} \le 1$, one can ensure that the halo is
bounded by the ghost layer, which significantly simplifies the communication
problem.
This assumption, however, defeats the purpose of using a semi-Lagrangian
approach, whose purpose is to enable large CFL values.

@@

To solve the communication problem, we first compute the adjacency matrix of the communication pattern, i.e.\ we construct the matrix $A_{P \times P}$, where $P$ is the number of processes, such that
\ben
a_{ij} = 
\left\{
\begin{array}{lr}
1 \hspace{5 mm} \text{if process `\textit{i}' sends a message to process `\textit{j}',} \\
0 \hspace{5 mm} \text{otherwise.}
\end{array}
\right.
\een
Note that this matrix is also distributed among processes, i.e.\ each row is owned by a separate rank. Next, we compute
\ben
S_i = \sum_j a_{ij} \hspace{5 mm} \text {and} \hspace{5 mm} R_i = \sum_j a_{ji} ,
\een
where $S_i$ and $R_i$ denote the number of messages sent and received, respectively. While $S_i$ can be computed trivially, a reduction operation is required to compute $R_i$. For instance, this can be achieved using a single \texttt{MPI\_Reduce\_scatter} function call. The last phase of the interpolation procedure involves overlapping the computation of interpolated values for local points with the communication of data between processes. This is done by alternating between local calculations and probing for incoming messages from other processes. The interpolation is finished once the values for all local the points have been calculated and all the remote requests have been processed (see Algorithm \ref{alg:interpolation}).
\begin{algorithm}[htbp]
\caption{$values \gets \texttt{Interpolate (}H, F, \underline{\mathbf{X}}\texttt{)}$: interpolate the value of $F$, defined on the local tree hierarchy $H$, at coordinates $\underline{\mathbf{X}}$}
\begin{algorithmic}[1]
\State $col \gets 0, \mathit{buff} \gets \mathit{null}$ \Comment{Phase I -- buffering}
\For {$\underline{\mathbf{p}} : \underline{\mathbf{X}}$} 
	\State $[r, \mathit{cell})] \gets H.\texttt{search(}\underline{\mathbf{p}}\texttt{)}$ \Comment{search for the owner's rank and cell}
	\If {$r = \mathit{mpirank}$}
        \State $\mathit{buff}[r].\texttt{push\_back(}\underline{\mathbf{p}}, cell\texttt{)}$
	\Else
        \State $\mathit{buff}[r].\texttt{push\_back(}\underline{\mathbf{p}}\texttt{)}$
        \State $\mathit{col}[r] \gets 1$
	\EndIf
\EndFor
\For {$r:\mathit{mpisize}$} \Comment{Phase II -- initiate communication and compute number of messages}
	\If {$\mathit{col}[r]$}
        \State $\texttt{MP\_Isend(}r, \mathit{buff}[r]\texttt{)}$
	\EndIf
\EndFor
\State  $S \gets \texttt{sum(}col\texttt{)}$ 
\State  $R \gets \texttt{MPI\_Reduce\_scatter(}\mathit{col},\texttt{MPI\_SUM)}$
\State $\mathit{done} \gets \mathit{false}, it \gets \mathit{buff}[mpirank].\texttt{begin()}$
\While {$\mathit{!done}$} \Comment{Phase III -- main loop}
	\If {$\mathit{it} \not= \mathit{buff[mpirank]}.\texttt{end()}$}
		\State $\mathit{values} \gets \texttt{process\_local(}it\texttt{)}$ \Comment{process local interpolations}
		\State $\texttt{++}\mathit{it}$
	\EndIf
	\If {$R > 0$} \Comment{process queries sent from remote processes}
		\State $[\mathit{msg, st}] \gets \texttt{MPI\_Iprobe()}$
		\If {$\mathit{msg}$}
			% \State $values \gets \texttt{MPI\_Recv(}st.\texttt{MPI\_SOURCE)}$ \Comment{}
			\State $\mathit{val\_buff} \gets \texttt{process\_queries(}\mathit{st}\texttt{)}$ \Comment{receive, search, and interpolate values}
			\State $\texttt{MPI\_Isend(}\mathit{st}.\texttt{MPI\_SOURCE},\mathit{val\_buff}\texttt{)}$ \Comment{send back interpolated values}
			\State $R\texttt{--}$
		\EndIf
	\EndIf
	\If {$S > 0$} \Comment{process replies sent to our queries}
		\State $\mathit{[msg, st]} \gets \texttt{MPI\_Iprobe()}$
		\If {$\mathit{msg}$}
			\State $\mathit{values} \gets \texttt{process\_replies(}\mathit{st}.\texttt{MPI\_SOURCE)}$ \Comment{receive remotely interpolated values}
			\State $S\texttt{--}$
		\EndIf
	\EndIf
	\State $\mathit{done} \gets S = 0 \: \And \: R = 0 \: \And \: \mathit{it = buff[mpirank]}.\texttt{end()}$
\EndWhile
\State \Return $\mathit{values}$
\end{algorithmic}
\label{alg:interpolation}
\end{algorithm}

Using the interpolation Algorithm \ref{alg:interpolation}, we close this section by presenting the final semi-Lagrangian Algorithm \ref{alg:semi-lagrangian}. The basic idea is to start from an initial guess $G^{n+1}_0$ for the grid and modify it using the refinement \eqref{eq:refine} and coarsening \eqref{eq:coarsen} criteria until convergence is obtained. Various options are available for $G^{n+1}_0$. For instance it is possible to start from the macromesh and only perform refinement steps until convergence. This choice, however, is not suitable since the first few iterations do not contain many cells and there is little work for parallelism. Here we simply take the previous grid as the starting point, i.e.\ $G^{n+1}_0 = G^n$. Note that this iterative process is essentially unavoidable since the grid is based on the values of the level-set function at $t^{n+1}$, which itself is unknown and is to be defined on $G^{n+1}$. Nonetheless the process converges to the final grid in at most $l_{\max}-l_{\min}$ steps where $l_{\min}$ and $l_{\max}$ denote the maximum and minimum depth of all trees in the forest, receptively.
\begin{algorithm}[htbp]
\caption{$[G^{n+1}, \phi^{n+1}] \gets \texttt{SemiLagrangian (}G^n, \phi^n, \underline{\mathbf{u}}^n, \underline{\mathbf{u}}^{n-1}, \text{CFL}\texttt{)}$: update $\phi^{n+1}$ from $\phi^n$ using a semi-Lagrangian scheme and construct the new forest $G^{n+1}$ that is consistent with the zero level-set of $\phi^{n+1}$}
\begin{algorithmic}[1]
\State $\Delta t_l \gets \text{CFL} \times G^n.\texttt{hmin()} / \max \{\underline{\mathbf{u}}^n \} $
\State $\Delta t   \gets \texttt{MPI\_Allreduce(}\Delta t_l,\texttt{MPI\_MIN)}$
\State $H^n \gets \texttt{Reconstruct(} G^n\texttt{)}$ 
\State $G^{n+1}_0 \gets G^n$
\While {$\mathit{true}$}
	\State $\underline{\mathbf{X}}_d  \gets \texttt{ComputeDeparturePoints(}G^{n+1}_0, \underline{\mathbf{u}}^n, \underline{\mathbf{u}}^{n-1}, \Delta t\texttt{)}$ \Comment{using equations \ref{eq:xstar} -- \ref{eq:vn_p_half}}
	\State $\phi^{n+1} \gets \texttt{Interpolate(}H^n, \phi^n, \underline{\textbf{X}}_d\texttt{)}$
	\State $G^{n+1} \gets G^{n+1}_0.\texttt{refine\_and\_coarsen(}\phi^{n+1}\texttt{)}$ \Comment{using equations \ref{eq:refine} and \ref{eq:coarsen} as criteria}
	\If {$G^{n+1} \not= G^{n+1}_0$}
		\State $G^{n+1}.\texttt{partition()}$
		\State $G^{n+1}_0 \gets G^{n+1}$
	\Else
		\State \textbf{break}
	\EndIf
\EndWhile

>>>>

Finally, in \texttt{p4est} trees are linearized, i.e.\ only the leaves are explicitly stored. However, explicit knowledge of the hierarchal structure of the tree is greatly beneficial in several algorithms, e.g. in search operations needed for the interpolation algorithm. Thus, we introduce a simple reconstruction algorithm that recreates a local representation of the entire ``forest'' that is only adapted to local cells and, potentially, the ghost layer. This approach is similar to the ideas introduced in \cite{Bangerth;Burstedde;Heister;etal:11:Algorithms-and-data-} and our tests show that in a typical application they amount to less that $1\%$ of the entire runtime. Algorithm \ref{alg:reconstruction} illustrates how this reconstruction is performed. Given a forest and a layer of ghost cells from \texttt{p4est}, the algorithm generates a local representation of the forest by recursively refining from the root until reaching the same level and location of all leaves in the local forest and the ghost layer. Note that algorithm \ref{alg:reconstruction} does not involve any communication and is load balanced provided that the initial forest is balanced. Figure \ref{fig:reconstruction} illustrates an example where Algorithm \ref{alg:reconstruction} is applied. Note how each process has independently generated a local representation of the forest that is refined to match the same leaves as in the global forest and ghost layer.
\begin{figure}[htbp]
\begin{center}
\includegraphics[width = \textwidth]{figures/reconstruct.pdf}
\end{center}
\caption{Left: a forest refined close to an interface and partitioned among four processes, as indicated by colors. Right: each process independently recreates a local forest that is refined to match the local grid and is as coarse as possible elsewhere. Note that empty cells are fictitious, i.e.\ they are only required to generate the hierarchal structure and are not matched by any corresponding cell in the global forest.}
\label{fig:reconstruction}
\end{figure}
\begin{algorithm}[htbp]
\caption{$H \gets \texttt{Reconstruct (}G\texttt{)}$: 
Construction of the local tree hierarchy, $H$, from the parallel grid, $G$, supplied by \texttt{p4est}. The algorithm starts at the coarsest level, i.e. the macromesh, and recursively splits the each cell to match the finest local and ghost cells generated by \texttt{p4est}.}
\begin{algorithmic}[1]
\State $H \gets G.\texttt{macromesh()}$ \Comment{start with the coarsest  grid}
\For {$\mathit{tr} \ein G.\texttt{local\_trees()}$} \Comment{update the hierarchy to match local cells}
	\For {$c \ein \mathit{tr}.\texttt{local\_cells()}$}
		\State $H.\texttt{update\_tree(} \mathit{tr}, c \texttt{)}$
	\EndFor
\EndFor

\For {$c \ein G.\texttt{ghost\_cells()}$} \Comment{update the hierarchy to mach local cells}
	\State $H.\texttt{update\_tree(} c.\texttt{get\_parent\_tree()}, c \texttt{)}$
\EndFor

\State \Return $H$
\\
\Function{$H.\texttt{update\_tree}$}{$\mathit{tr}, c$}   \Comment{recursive tree reconstruction}
	\State $c_l \gets H.\texttt{get\_root\_cell(}\mathit{tr}\texttt{)}$ 
	\While{$c_l.\texttt{level()} \not= c.\texttt{level()} $} \Comment{recursively search for a cell of the same size as \textit{c}}
		\If {$c_l.$\texttt{is\_leaf()}} $c_l$\texttt{.split()} \Comment{if the current cell is leaf, split the cell and continue search}
		\EndIf
		\State $h \gets c_l.\texttt{length()}$ \Comment{select the search path based on cell coordinate.}
		\State $i \gets c.x \ge c_l.x + h/2$	
		\State $j \gets c.y \ge c_l.y + h/2$
		\State $k \gets c.z \ge c_l.z + h/2$

		\State $c_l \gets c_l.\texttt{get\_child(} i, j, k \texttt{)}$
	\EndWhile	
\EndFunction
\end{algorithmic}
\label{alg:reconstruction}
\end{algorithm}

\subsection{Interpolation and semi-Lagrangian methods} \label{sec:interpolation_semilagrangian}
As indicated earlier, we use the semi-Lagrangian method to solve equation \eqref{eq:ls} when the velocity field is externally generated, i.e.\ when it does not depend explicitly on the level-set function itself. Let us rewrite equation \eqref{eq:ls} along the characteristic curve $\underline{\mathbf{X}}(t)$ as:
\be
\left\{
\begin{array}{rcl}
\dfrac{\text{d} \underline{\mathbf{X}}}{\text{d} t} &=& \underline{\mbf{u}}, \\ [3ex]
\dfrac{\text{d} \phi(\underline{\mathbf{X}}(t), t)}{\text{d} t} &=& 0.
\end{array}
\right.
\label{eq:sl}
\ee
The semi-Lagrangian method integrates equations \eqref{eq:sl} backward in time,
i.e.\ starting from the grid $G^{n+1}$ (computed in iteratively as explained
later on), we simply write $\phi^{n+1}(\underline{\mathbf{X}}^{n+1}) =
\phi(\underline{\mathbf{X}}(t^{n+1}), t^{n+1}) =
\phi(\underline{\mathbf{X}}(t^n), t^n) = \phi^n(\underline{\mathbf{X}}_d)$.
Here, the characteristic curves are chosen such that $\underline{\mathbf{X}}(t^{n+1})$ are simply the coordinates of grids points of $G^{n+1}$, and $\underline{\mathbf{X}}_d$ are the departure points, which are computed using the second-order midpoint method \cite{Min;Gibou:07:A-second-order-accur}:
\bea
\underline{\mathbf{X}}^\star &=& \underline{\mathbf{X}}^{n+1} - \frac{\Delta t}{2} \underline{\mbf{u}}^{n}(\underline{\mathbf{X}}^n),	   \label{eq:xstar}      \\
\underline{\mathbf{X}}_d     &=& \underline{\mathbf{X}}^{n+1} - \Delta t \underline{\mbf{u}}^{n+\frac{1}{2}}(\underline{\mathbf{X}}^\star), \label{eq:xdeparture}
\eea
where $\underline{\mbf{u}}^{n+\frac{1}{2}}$ is obtained via extrapolation from previous times, i.e.:
\be
\underline{\mbf{u}}^{n+\frac{1}{2}} = \frac{3}{2} \underline{\mbf{u}}^n - \frac{1}{2}\underline{\mbf{u}}^{n-1}. \label{eq:vn_p_half}
\ee
Note that all values at the intermediate point, $\underline{\mathbf{X}}^\star$,
and departure point, $\underline{\mathbf{X}}_d$, must be calculated via
interpolation from the previous grids $G^{n}$ and $G^{n-1}$.
Here, we use the stabilized second-order interpolation for
$\phi(\underline{\mathbf{X}}_d)$ and the multi-linear interpolation for
$\underline{\mbf{u}}^{n+\frac{1}{2}}(\underline{\mathbf{X}}^\star)$
\cite{Min;Gibou:07:A-second-order-accur}.
Although parallelization of the interpolation process on a shared-memory
machine is trivial, the same cannot be said for distributed-memory machines.
In fact, the parallel interpolation given in Algorithm \ref{alg:interpolation}
is probably the most important contribution of this article since this
procedure, which is trivial on uniform grids, is challenging in the case of
trees because it is not straightforward to identify which processes owns the
departure points.
Indeed, complications arise because not all departure points will reside in the
domain owned by the current process. Moreover, due to the irregular shapes of
the partitions, one cannot even ensure they are entirely owned by neighboring
processes.
At best we can only expect that their locations are bounded by a halo of width
$w \le \text{CFL} \: \Delta x_{\min}$ around the local partition, where
$x_{\min}$ is the size of the smallest cell in the forest.
Naturally, if one enforces $\text{CFL} \le 1$, one can ensure that the halo is
bounded by the ghost layer, which significantly simplifies the communication
problem.
This assumption, however, defeats the purpose of using a semi-Lagrangian
approach, whose purpose is to enable large CFL values.

@@

To solve the communication problem, we first compute the adjacency matrix of the communication pattern, i.e.\ we construct the matrix $A_{P \times P}$, where $P$ is the number of processes, such that
\ben
a_{ij} = 
\left\{
\begin{array}{lr}
1 \hspace{5 mm} \text{if process `\textit{i}' sends a message to process `\textit{j}',} \\
0 \hspace{5 mm} \text{otherwise.}
\end{array}
\right.
\een
Note that this matrix is also distributed among processes, i.e.\ each row is owned by a separate rank. Next, we compute
\ben
S_i = \sum_j a_{ij} \hspace{5 mm} \text {and} \hspace{5 mm} R_i = \sum_j a_{ji} ,
\een
where $S_i$ and $R_i$ denote the number of messages sent and received, respectively. While $S_i$ can be computed trivially, a reduction operation is required to compute $R_i$. For instance, this can be achieved using a single \texttt{MPI\_Reduce\_scatter} function call. The last phase of the interpolation procedure involves overlapping the computation of interpolated values for local points with the communication of data between processes. This is done by alternating between local calculations and probing for incoming messages from other processes. The interpolation is finished once the values for all local the points have been calculated and all the remote requests have been processed (see Algorithm \ref{alg:interpolation}).
\begin{algorithm}[htbp]
\caption{$values \gets \texttt{Interpolate (}H, F, \underline{\mathbf{X}}\texttt{)}$: interpolate the value of $F$, defined on the local tree hierarchy $H$, at coordinates $\underline{\mathbf{X}}$. This is achieved in three phases as detailed in the text. I) First, interpolation points are buffered in two arrays, those than can be interpolated locally and those that should be sent to other processes. II) In the second phase, each process initiates the remote buffer exchange and computes the number of messages it should expect to receive using the communication matrix idea described in section \ref{sec:interpolation_semilagrangian}. III) Finally, each process alternates between computing local interpolation and receiving and performing remote interpolation requests from other processes.}
\begin{algorithmic}[1]
\State $col \gets 0, \mathit{local\_buff} \gets \mathit{null}, \mathit{remote\_buff} \gets \mathit{null}$ \Comment{Phase I -- buffering}
\For {$\underline{\mathbf{p}} \ein \underline{\mathbf{X}}$} 
	\State $[owners\_rank, \mathit{cell}] \gets H.\texttt{search(}\underline{\mathbf{p}}\texttt{)}$ \Comment{search for the owner's rank and cell}
	\If {$owners\_rank = \mathit{mpirank}$} \Comment{can interpolate locally}
        \State $\mathit{local\_buff}[owners\_rank].\texttt{push\_back(}\underline{\mathbf{p}}, cell\texttt{)}$ 
	\Else \Comment{requires remote interpolation}
        \State $\mathit{remote\_buff}[owners\_rank].\texttt{push\_back(}\underline{\mathbf{p}}\texttt{)}$ 
        \State $\mathit{col}[owners\_rank] \gets 1$
	\EndIf
\EndFor
\\
\For {$r \ein \mathit{[0,mpisize)}$} \Comment{Phase II -- initiate communication and compute number of messages}
	\If {$\mathit{col}[r] \deq 1$} 
        \State $req \gets \texttt{MP\_Isend(}r, \texttt{QUERY\_TAG}, \mathit{remote\_buff}[r]\texttt{)}$ \Comment{initiate a non-blocking send to process \textit{r}}
        	\State $\textit{query\_requests}.\texttt{push\_back(}req\texttt{)}$
	\EndIf
\EndFor
\State  $S \gets \texttt{sum(}col\texttt{)}$ \Comment{compute number of messages to send}
\State  $R \gets \texttt{MPI\_Reduce\_scatter(}\mathit{col},\texttt{MPI\_SUM)}$ \Comment{compute number of messages to receive}
\\
\State $\mathit{done} \gets \mathit{false}$ \Comment{Phase III -- interpolation}
\State $it \gets \mathit{local\_buff}[mpirank].\texttt{begin()}$ 
\While {$\mathit{!done}$} 
	\If {$\mathit{it} \not= \mathit{buff[mpirank]}.\texttt{end()}$}
		\State $\mathit{values} \gets \texttt{process\_local\_interpolation(}it\texttt{)}$ \Comment{process local interpolations}
		\State $\texttt{++}\mathit{it}$
	\EndIf
	\If {$R > 0$} \Comment{search for interpolation query in the message queue}
		\State $\mathit{message} \gets \texttt{MPI\_Iprobe(\texttt{MPI\_ANY\_SOURCE}, \texttt{QUERY\_TAG})}$
		\If {$\mathit{message}.\texttt{is\_pending()}$}
			\State $\mathit{values} \gets \texttt{process\_remote\_queries(}\mathit{message}\texttt{)}$ \Comment{receive, search, and interpolate values}
			\State $req \gets \texttt{MPI\_Isend(}\mathit{message}.\texttt{MPI\_SOURCE}, \texttt{REPLY\_TAG}, \mathit{values}\texttt{)}$ \Comment{send back interpolated values}
			\State $\textit{reply\_requests}.\texttt{push\_back(}req\texttt{)}$
			\State $R\texttt{--}$
		\EndIf
	\EndIf
	\If {$S > 0$} \Comment{search for interpolation reply in the message queue}
		\State $\mathit{message} \gets \texttt{MPI\_Iprobe(\texttt{MPI\_ANY\_SOURCE}, \texttt{REPLY\_TAG})}$
		\If {$\mathit{message}.\texttt{is\_pending()}$}
			\State $\mathit{values} \gets \texttt{process\_replies(}\mathit{message}\texttt{)}$ \Comment{receive remotely interpolated values}
			\State $S\texttt{--}$
		\EndIf
	\EndIf
	\State $\mathit{done} \gets S \deq 0 \And R \deq 0 \And \mathit{it \deq local\_buff[mpirank]}.\texttt{end()}$
\EndWhile
\State $\texttt{MPI\_Waitall(}query\_requests, reply\_requests\texttt{)}$ \Comment{make sure all messages have been received}
\State \Return $\mathit{values}$
\end{algorithmic}
\label{alg:interpolation}
\end{algorithm}

Using the interpolation Algorithm \ref{alg:interpolation}, we close this section by presenting the final semi-Lagrangian Algorithm \ref{alg:semi-lagrangian}. The basic idea is to start from an initial guess $G^{n+1}_0$ for the grid and modify it using the refinement \eqref{eq:refine} and coarsening \eqref{eq:coarsen} criteria until convergence is obtained. Various options are available for $G^{n+1}_0$. For instance it is possible to start from the macromesh and only perform refinement steps until convergence. This choice, however, is not suitable since the first few iterations do not contain many cells and there is little work for parallelism. Here we simply take the previous grid as the starting point, i.e.\ $G^{n+1}_0 = G^n$. Note that this iterative process is essentially unavoidable since the grid is based on the values of the level-set function at $t^{n+1}$, which itself is unknown and is to be defined on $G^{n+1}$. Nonetheless the process converges to the final grid in at most $l_{\max}-l_{\min}$ steps where $l_{\min}$ and $l_{\max}$ denote the maximum and minimum depth of all trees in the forest, receptively.
\begin{algorithm}[htbp]
\caption{$[G^{n+1}, \phi^{n+1}] \gets \texttt{SemiLagrangian (}G^n, \phi^n, \underline{\mathbf{u}}^n, \underline{\mathbf{u}}^{n-1}, \text{CFL}\texttt{)}$: update $\phi^{n+1}$ from $\phi^n$ using a semi-Lagrangian scheme and construct the new forest $G^{n+1}$ that is consistent with the zero level-set of $\phi^{n+1}$}
\begin{algorithmic}[1]
\State $\Delta t_l \gets \text{CFL} \times G^n.\texttt{hmin()} / \max \{\underline{\mathbf{u}}^n \} $
\State $\Delta t   \gets \texttt{MPI\_Allreduce(}\Delta t_l,\texttt{MPI\_MIN)}$ \Comment{compute $\Delta t$ based on CFL condition across all processes}
\State $H^n \gets \texttt{Reconstruct(} G^n\texttt{)}$  \Comment{using algorithm \ref{alg:reconstruction}}
\State $G^{n+1}_0 \gets G^n$
\While {$\mathit{true}$}
	\State $\underline{\mathbf{X}}_d  \gets \texttt{ComputeDeparturePoints(}G^{n+1}_0, \underline{\mathbf{u}}^n, \underline{\mathbf{u}}^{n-1}, \Delta t\texttt{)}$ \Comment{using equations \ref{eq:xstar} -- \ref{eq:vn_p_half}}
	\State $\phi^{n+1} \gets \texttt{Interpolate(}H^n, \phi^n, \underline{\textbf{X}}_d\texttt{)}$ \Comment{using algorithm \ref{alg:interpolation}}
	\State $G^{n+1} \gets G^{n+1}_0.\texttt{refine\_and\_coarsen(}\phi^{n+1}\texttt{)}$ \Comment{using equations \ref{eq:refine} and \ref{eq:coarsen} as criteria}
	\If {$G^{n+1} \not= G^{n+1}_0$}
		\State $G^{n+1}.\texttt{partition()}$
		\State $G^{n+1}_0 \gets G^{n+1}$
	\Else
		\State \textbf{break}
	\EndIf
\EndWhile
